---
title: "Cyclistic User Behavior Analysis (Monthly Data)"
author: "Junior Data Analyst"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    keep_md: false
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE # Suppress all warnings globally
  )

#Install necessary packages
if (!requireNamespace("data.table", quietly = TRUE)) install.packages("data.table")
if (!requireNamespace("mapdeck", quietly = TRUE)) install.packages("mapdeck")
if (!requireNamespace("plotly", quietly = TRUE)) install.packages("plotly")
if (!requireNamespace("arrow", quietly = TRUE)) install.packages("arrow")
if (!requireNamespace("leaflet", quietly = TRUE)) install.packages("leaflet")
if (!requireNamespace("kableExtra", quietly = TRUE)) install.packages("kableExtra")

# Load necessary libraries
library(data.table)  # Efficient data manipulation and date/time handling
library(mapdeck)     # Scalable and high-performance mapping
library(plotly)      # Interactive visualizations of large data sets
library(arrow)       # Optimized storage and processing of large data sets
library(leaflet)     # Interactive visualizations of small data sets
library(kableExtra)  # Load kableExtra to style tables
```

## 2. Data Preparation
### Load and Combine Monthly Data
```{r}
# List of monthly data file paths
file_paths <- list(
  "divvy-tripdata/202312-divvy-tripdata.csv",
  "divvy-tripdata/202401-divvy-tripdata.csv",
  "divvy-tripdata/202402-divvy-tripdata.csv",
  "divvy-tripdata/202403-divvy-tripdata.csv",
  "divvy-tripdata/202404-divvy-tripdata.csv",
  "divvy-tripdata/202405-divvy-tripdata.csv",
  "divvy-tripdata/202406-divvy-tripdata.csv",
  "divvy-tripdata/202407-divvy-tripdata.csv",
  "divvy-tripdata/202408-divvy-tripdata.csv",
  "divvy-tripdata/202409-divvy-tripdata.csv",
  "divvy-tripdata/202410-divvy-tripdata.csv",
  "divvy-tripdata/202411-divvy-tripdata.csv"
)

# Load and combine data using data.table
monthly_data <- rbindlist(lapply(file_paths, fread))
```

### 2.1 Check and Clean the Data
#### Check for Missing Values
```{r}
# Check for missing values
missing_summary <- colSums(is.na(monthly_data))
```

#### Remove Missing Values
```{r}
# Remove rows with missing values
monthly_data <- monthly_data[complete.cases(monthly_data)]
```

#### Ensure Proper Date/Time Format
```{r}
# Convert date/time fields to proper format
monthly_data[, started_at := as.POSIXct(started_at, format = "%Y-%m-%d %H:%M:%S")]
monthly_data[, ended_at := as.POSIXct(ended_at, format = "%Y-%m-%d %H:%M:%S")]
```

#### Remove Duplicate Records
```{r}
# Remove duplicate rows
monthly_data <- unique(monthly_data)
```

### 2.2 Add New Calculated Columns
#### Trip Duration
```{r}
# Calculate trip duration in minutes
monthly_data[, trip_duration := as.numeric(difftime(ended_at, started_at, units = "mins"))]
```

#### Day of Week
```{r}
# Extract the day of the week
monthly_data[, day_of_week := weekdays(started_at)]
```

#### Hour of the Day
```{r}
# Extract the hour from the start time
monthly_data[, hour_of_day := as.numeric(format(started_at, "%H"))]
```

## 3. Analyze the Data
### 3.1 Aggregate Metrics by User Type
#### Total Number of Rides by User Type
```{r}
# Total number of rides by user type
total_rides <- monthly_data[, .N, by = member_casual]


# Styled table with kableExtra
total_rides %>%
  knitr::kable(
    col.names = c("User Type", "Total Rides"),
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )
```

#### Average Trip Duration by User Type
```{r}
# Average trip duration by user type
avg_duration <- monthly_data[, .(avg_duration = mean(trip_duration, na.rm = TRUE)), by = member_casual]

# Average trip duration table
avg_duration %>%
  knitr::kable(
    col.names = c("User Type", "Average Duration (minutes)"),
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE
  )


# Interactive Plot
plot_ly(avg_duration, x = ~member_casual, y = ~avg_duration, type = 'bar', color = ~member_casual) %>%
  layout(title = "Average Trip Duration by User Type",
         xaxis = list(title = "User Type"),
         yaxis = list(title = "Average Duration (minutes)"))
```

### 3.2 Analyze Usage Patterns
#### Popular Days by User Type
```{r}
# Count rides by day of the week and user type
popular_days <- monthly_data[, .N, by = .(member_casual, day_of_week)]
popular_days$day_of_week <- factor(popular_days$day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))
setnames(popular_days, "N", "TotalRides")


# Separate tables for each user type
popular_days_member <- popular_days[member_casual == "member"][order(-TotalRides)]
popular_days_casual <- popular_days[member_casual == "casual"][order(-TotalRides)]

# Member Table
member_table <- popular_days_member %>%
  knitr::kable(
    col.names = c("User Type", "Day of Week", "Total Rides"),
    caption = "Popular Days of the Week for Members",
    format = "html"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 14
  ) %>%
  column_spec(1, bold = TRUE, color = "white", background = "steelblue") %>% # Style User Type column
  column_spec(2, width = "20em") %>% # Adjust column width for Day of Week
  column_spec(3, color = "black", background = "lightgray") # Style Total Rides column

# Casual Table
casual_table <- popular_days_casual %>%
  knitr::kable(
    col.names = c("User Type", "Day of Week", "Total Rides"),
    caption = "Popular Days of the Week for Casual Riders",
    format = "html"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 14
  ) %>%
  column_spec(1, bold = TRUE, color = "white", background = "red") %>% # Style User Type column
  column_spec(2, width = "20em") %>% # Adjust column width for Day of Week
  column_spec(3, color = "black", background = "lightgray") # Style Total Rides column

# Print tables
member_table
casual_table


# Interactive Plot
plot_ly(popular_days, x = ~day_of_week, y = ~TotalRides, color = ~member_casual, type = 'bar') %>%
  layout(title = "Rides by Day of the Week and User Type",
         xaxis = list(title = "Day of the Week"),
         yaxis = list(title = "Number of Rides"))
```

#### Popular Months by User Type
```{r}
# Extract month from started_at
monthly_data[, month := format(started_at, "%B")]

# Count rides by month and user type
popular_months <- monthly_data[, .N, by = .(member_casual, month)]
popular_months$month <- factor(popular_months$month, levels = month.name)
setnames(popular_months, "N", "TotalRides")

# Separate tables for each user type and sort by Total Rides
popular_months_member <- popular_months[member_casual == "member"][order(-TotalRides)]
popular_months_casual <- popular_months[member_casual == "casual"][order(-TotalRides)]

# Member Table
member_month_table <- popular_months_member %>%
  knitr::kable(
    col.names = c("User Type", "Month", "Total Rides"),
    caption = "Popular Months for Members",
    format = "html"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 14
  ) %>%
  column_spec(1, bold = TRUE, color = "white", background = "steelblue") %>% # Style User Type column
  column_spec(2, width = "20em") %>% # Adjust column width for Month
  column_spec(3, color = "black", background = "lightgray") # Style Total Rides column

# Casual Table
casual_month_table <- popular_months_casual %>%
  knitr::kable(
    col.names = c("User Type", "Month", "Total Rides"),
    caption = "Popular Months for Casual Riders",
    format = "html"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 14
  ) %>%
  column_spec(1, bold = TRUE, color = "white", background = "red") %>% # Style User Type column
  column_spec(2, width = "20em") %>% # Adjust column width for Month
  column_spec(3, color = "black", background = "lightgray") # Style Total Rides column

# Print tables
member_month_table
casual_month_table

# Interactive Plot
plot_ly(popular_months, x = ~month, y = ~TotalRides, color = ~member_casual, type = 'bar') %>%
  layout(title = "Rides by Month and User Type",
         xaxis = list(title = "Month"),
         yaxis = list(title = "Number of Rides"))
```

#### Popular Start Station of the Month by User Type
```{r}
# Step 1: Count rows with missing or blank start_station_name
missing_stations <- monthly_data[is.na(start_station_name) | start_station_name == "", .N]

# Step 2: Filter out rows with missing or blank start_station_name
monthly_data <- monthly_data[!(is.na(start_station_name) | start_station_name == "")]

# Step 3: Count rides by start station, month, and user type
popular_start_stations <- monthly_data[, .N, by = .(start_station_name, month, member_casual)]
setnames(popular_start_stations, "N", "TotalRides") # Rename column for clarity
popular_start_stations <- popular_start_stations[order(-TotalRides)] # Order by Total Rides descending

# Step 4: Filter top start stations for each month and user type
top_start_stations <- popular_start_stations[, .SD[which.max(TotalRides)], by = .(month, member_casual)]

# Step 5: Adjust text placement for visibility in visualizations
top_start_stations[, text_position := ifelse(TotalRides < 50, "inside", "outside")]

# Remove the `text_position` column from the data
top_start_stations2 <- top_start_stations[, .(start_station_name, month, member_casual, TotalRides)]

# Separate tables for each user type
top_start_stations_member <- top_start_stations2[member_casual == "member"][order(-TotalRides)]
top_start_stations_casual <- top_start_stations2[member_casual == "casual"][order(-TotalRides)]

# Member Table
member_table <- top_start_stations_member %>%
  knitr::kable(
    col.names = c("Start Station", "Month", "User Type", "Total Rides"),
    caption = "Top Start Stations for Members",
    format = "html"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 14
  ) %>%
  column_spec(1, bold = TRUE, color = "white", background = "steelblue") %>% # Style Start Station column
  column_spec(3, width = "10em") %>% # Adjust User Type column width
  column_spec(4, color = "black", background = "lightgray") # Style Total Rides column

# Casual Table
casual_table <- top_start_stations_casual %>%
  knitr::kable(
    col.names = c("Start Station", "Month", "User Type", "Total Rides"),
    caption = "Top Start Stations for Casual Riders",
    format = "html"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 14
  ) %>%
  column_spec(1, bold = TRUE, color = "white", background = "red") %>% # Style Start Station column
  column_spec(3, width = "10em") %>% # Adjust User Type column width
  column_spec(4, color = "black", background = "lightgray") # Style Total Rides column

# Print tables
member_table
casual_table


# Interactive Plot with Station Name Annotations
plot_ly(top_start_stations, 
        x = ~month, 
        y = ~TotalRides, 
        color = ~member_casual, 
        type = 'bar', 
        text = ~paste("Station: ", start_station_name), 
        textposition = ~text_position, 
        hoverinfo = "text+y") %>%
  layout(title = "Top Start Stations by Month and User Type",
         xaxis = list(title = "Month", categoryorder = "array", categoryarray = month.name),
         yaxis = list(title = "Number of Rides"),
         barmode = "group",
         showlegend = TRUE)

```

#### Popular End Station of the Month by User Type
```{r}
# Count rows with missing or blank end_station_name
missing_stations <- monthly_data[is.na(end_station_name) | end_station_name == "", .N]

# Filter out rows with missing or blank end_station_name
monthly_data <- monthly_data[!(is.na(end_station_name) | end_station_name == "")]

# Count rides by end station, month, and user type
popular_end_stations <- monthly_data[, .N, by = .(end_station_name, month, member_casual)]
setnames(popular_end_stations, "N", "TotalRides") # Rename column for clarity
popular_end_stations <- popular_end_stations[order(-TotalRides)] # Order by Total Rides descending

# Filter top end stations for each month and user type
top_end_stations <- popular_end_stations[, .SD[which.max(TotalRides)], by = .(month, member_casual)]

# Adjust text placement to ensure visibility
top_end_stations[, text_position := ifelse(TotalRides < 50, "inside", "outside")]

# Remove the `text_position` column from the data
top_end_stations2 <- top_end_stations[, .(end_station_name, month, member_casual, TotalRides)]

# Separate tables for each user type
top_end_stations_member <- top_end_stations2[member_casual == "member"][order(-TotalRides)]
top_end_stations_casual <- top_end_stations2[member_casual == "casual"][order(-TotalRides)]

# Member Table
member_end_station_table <- top_end_stations_member %>%
  knitr::kable(
    col.names = c("End Station", "Month", "User Type", "Total Rides"),
    caption = "Top End Stations for Members",
    format = "html"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 14
  ) %>%
  column_spec(1, bold = TRUE, color = "white", background = "steelblue") %>% # Style End Station column
  column_spec(3, width = "10em") %>% # Adjust User Type column width
  column_spec(4, color = "black", background = "lightgray") # Style Total Rides column

# Casual Table
casual_end_station_table <- top_end_stations_casual %>%
  knitr::kable(
    col.names = c("End Station", "Month", "User Type", "Total Rides"),
    caption = "Top End Stations for Casual Riders",
    format = "html"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 14
  ) %>%
  column_spec(1, bold = TRUE, color = "white", background = "red") %>% # Style End Station column
  column_spec(3, width = "10em") %>% # Adjust User Type column width
  column_spec(4, color = "black", background = "lightgray") # Style Total Rides column

# Print tables
member_end_station_table
casual_end_station_table
# Interactive Plot with Station Name Annotations
plot_ly(top_end_stations, 
        x = ~month, 
        y = ~TotalRides, 
        color = ~member_casual, 
        type = 'bar', 
        text = ~paste("Station: ", end_station_name), 
        textposition = ~text_position, 
        hoverinfo = "text+y") %>%
  layout(title = "Top End Stations by Month and User Type",
         xaxis = list(title = "Month", categoryorder = "array", categoryarray = month.name),
         yaxis = list(title = "Number of Rides"),
         barmode = "group",
         showlegend = TRUE)


```


#### Peak Hours by User Type
```{r}
# Aggregate data for peak hours and days
heatmap_data <- monthly_data[, .(Total_Rides = .N), by = .(hour_of_day, day_of_week, member_casual)]
heatmap_data[, day_of_week := factor(day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))]

# Calculate average total rides for each hour across 12 months
peak_hours <- monthly_data[, .(Avg_Total_Rides = mean(.N, na.rm = TRUE)), 
                           by = .(hour_of_day, member_casual)]

# Rename column for clarity
setnames(peak_hours, "Avg_Total_Rides", "TotalRides")

# Sort data by Total Rides in descending order
peak_hours <- peak_hours[order(-TotalRides)]

# Separate tables for each user type
peak_hours_member <- peak_hours[member_casual == "member"][1:12]
peak_hours_casual <- peak_hours[member_casual == "casual"][1:12]

# Member Table
member_peak_hours_table <- peak_hours_member %>%
  knitr::kable(
    col.names = c("Hour of Day", "User Type", "Total Rides"),
    caption = "Peak Hours of the Day for Members (Average Rides)",
    format = "html"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 14
  ) %>%
  column_spec(1, bold = TRUE, color = "white", background = "steelblue") %>% # Style Hour of Day column
  column_spec(3, color = "black", background = "lightgray") # Style Total Rides column

# Casual Table
casual_peak_hours_table <- peak_hours_casual %>%
  knitr::kable(
    col.names = c("Hour of Day", "User Type", "Total Rides"),
    caption = "Peak Hours of the Day for Casual Riders (Average Rides)",
    format = "html"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed"),
    full_width = FALSE,
    font_size = 14
  ) %>%
  column_spec(1, bold = TRUE, color = "white", background = "red") %>% # Style Hour of Day column
  column_spec(3, color = "black", background = "lightgray") # Style Total Rides column

# Print tables
member_peak_hours_table
casual_peak_hours_table

# Aggregate data for peak hours
hourly_data <- monthly_data[, .N, by = .(hour_of_day, member_casual)]

# Create a bar graph with plot_ly
plot_ly(
  data = hourly_data,
  x = ~hour_of_day,
  y = ~N,
  color = ~member_casual,
  type = "bar",
  hoverinfo = "x+y+name",
  barmode = "group"
) %>%
  layout(
    title = "Peak Hours by User Type",
    xaxis = list(title = "Hour of Day", tickmode = "array", tickvals = 0:23),
    yaxis = list(title = "Number of Rides"),
    legend = list(title = list(text = "User Type")),
    margin = list(t = 50, b = 50, l = 50, r = 50)
  )


# Create heatmap for member riders
member_heatmap <- heatmap_data[member_casual == "member"]
member_plot <- plot_ly(
  data = member_heatmap,
  x = ~hour_of_day,
  y = ~day_of_week,
  z = ~Total_Rides,
  type = "heatmap",
  colorscale = list(c(0, "lightblue"), c(1, "darkblue")),
  hoverinfo = "x+y+z",
  showscale = TRUE
) %>%
  layout(
    xaxis = list(title = "Hour of Day"),
    yaxis = list(title = "Day of Week"),
    colorbar = list(title = "Total Rides")
  )

# Create heatmap for casual riders
casual_heatmap <- heatmap_data[member_casual == "casual"]
casual_plot <- plot_ly(
  data = casual_heatmap,
  x = ~hour_of_day,
  y = ~day_of_week,
  z = ~Total_Rides,
  type = "heatmap",
  colorscale = list(c(0, "lightpink"), c(1, "red")),
  hoverinfo = "x+y+z",
  showscale = TRUE
) %>%
  layout(
    xaxis = list(title = "Hour of Day"),
    yaxis = list(title = "Day of Week"),
    colorbar = list(title = "Total Rides")
  )

# Combine the plots with labels
subplot(member_plot, casual_plot, nrows = 2, shareX = TRUE, titleX = TRUE, titleY = TRUE) %>%
  layout(
    title = "Peak Hours by User Type",
    annotations = list(
      list(
        x = 0.5,
        y = 1,
        text = "Member Riders",
        showarrow = FALSE,
        xref = "paper",
        yref = "paper",
        font = list(size = 14, color = "yellow")
      ),
      list(
        x = 0.5,
        y = 0.5,
        text = "Casual Riders",
        showarrow = FALSE,
        xref = "paper",
        yref = "paper",
        font = list(size = 14, color = "red")
      )
    )
  )


```

## 4. Visualize Insights

### Popular Start Station of the Month by User Type
```{r}

# Count rows with missing or blank start_station_name
missing_stations <- monthly_data[is.na(start_station_name) | start_station_name == "", .N]

# Filter out rows with missing or blank start_station_name
monthly_data <- monthly_data[!(is.na(start_station_name) | start_station_name == "")]

# Count rides by start station, month, and user type, including latitude and longitude
popular_start_stations <- monthly_data[, .(
  total_rides = .N,
  start_lat = mean(start_lat, na.rm = TRUE),
  start_lng = mean(start_lng, na.rm = TRUE)
), by = .(start_station_name, month, member_casual)]

# Filter top start stations for each month and user type
top_start_stations <- popular_start_stations[, .SD[which.max(total_rides)], by = .(month, member_casual)]

# Adjust text placement for annotations
top_start_stations[, text_position := ifelse(total_rides < 50, "inside", "outside")]

# Create tooltip column in popular_start_stations
popular_start_stations[, tooltip := paste(
  "<b>Station Name:</b>", start_station_name, "<br>",
  "<b>Total Rides:</b>", total_rides, "<br>",
  "<b>Month:</b>", month, "<br>",
  "<b>User Type:</b>", member_casual
)]

# Map: Popular start stations by month with legend
mapdeck(
  token = "your_mapbox_token",
  style = mapdeck_style("dark"),
  location = c(mean(popular_start_stations$start_lng, na.rm = TRUE), 
               mean(popular_start_stations$start_lat, na.rm = TRUE)),
  zoom = 10
) %>%
  add_scatterplot(
    data = popular_start_stations,
    lat = "start_lat", lon = "start_lng", radius = "total_rides",
    fill_colour = "member_casual", # Color points by user type
    layer_id = "popular_start_stations",
    tooltip = "tooltip",
    legend = TRUE, # Enable legend
    legend_options = list(
      title = "User Type", 
      style = list(fontSize = "14px")
    ),
    update_view = FALSE # Keep the map centered
  )



```

### Map of popular start stations using leaflet
```{r}
# Filter to top stations for each month and user type
top_stations <- popular_start_stations[, .SD[which.max(total_rides)], by = .(month, member_casual)]

# Create leaflet map
leaflet(data = top_stations) %>%
  addTiles() %>% # Add base map tiles
  addCircleMarkers(
    lat = ~start_lat,
    lng = ~start_lng,
    radius = ~sqrt(total_rides) / 2, # Scale marker size by number of rides
    color = ~ifelse(member_casual == "member", "blue", "red"), # Color by user type
    group = ~member_casual, # Group for filtering
    label = ~paste(
      "Station Name: ", start_station_name, " ",
      "Month: ", month, " ",
      "Total Rides: ", total_rides
    ), # Hover information
    labelOptions = labelOptions(
      style = list(
        "font-weight" = "bold", 
        "font-size" = "12px"
      ),
      textsize = "15px",
      direction = "auto"
    ),
    popup = ~paste0(
      "<b>Station Name:</b> ", start_station_name, "<br>",
      "<b>Total Rides:</b> ", total_rides, "<br>",
      "<b>Month:</b> ", month, "<br>",
      "<b>User Type:</b> ", member_casual
    ) # Clickable information
  ) %>%
  addLayersControl(
    overlayGroups = c("member", "casual"), # Enable filtering by user type
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addLegend(
    position = "bottomright",
    colors = c("blue", "red"),
    labels = c("Member Riders", "Casual Riders"),
    title = "User Type"
  )

```

### Popular End Station of the Month by User Type
```{r}

# Count rows with missing or blank end_station_name
missing_stations <- monthly_data[is.na(end_station_name) | end_station_name == "", .N]

# Filter out rows with missing or blank end_station_name
monthly_data <- monthly_data[!(is.na(end_station_name) | end_station_name == "")]

# Count rides by end station, month, and user type, including latitude and longitude
popular_end_stations <- monthly_data[, .(
  total_rides = .N,
  end_lat = mean(end_lat, na.rm = TRUE),
  end_lng = mean(end_lng, na.rm = TRUE)
), by = .(end_station_name, month, member_casual)]

# Filter top end stations for each month and user type
top_end_stations <- popular_end_stations[, .SD[which.max(total_rides)], by = .(month, member_casual)]

# Adjust text placement for annotations
top_end_stations[, text_position := ifelse(total_rides < 50, "inside", "outside")]

# Create tooltip column in popular_end_stations
top_end_stations[, tooltip := paste(
  "<b>Station Name:</b>", end_station_name, "<br>",
  "<b>Total Rides:</b>", total_rides, "<br>",
  "<b>Month:</b>", month, "<br>",
  "<b>User Type:</b>", member_casual
)]

# Map: Popular end stations by month with legend
mapdeck(
  token = "your_mapbox_token",
  style = mapdeck_style("dark"),
  location = c(mean(popular_end_stations$end_lng, na.rm = TRUE), 
               mean(popular_end_stations$end_lat, na.rm = TRUE)),
  zoom = 10
) %>%
  add_scatterplot(
    data = top_end_stations,
    lat = "end_lat", lon = "end_lng", radius = "total_rides",
    fill_colour = "member_casual", # Color points by user type
    layer_id = "popular_end_stations",
    tooltip = "tooltip",
    legend = TRUE, # Enable legend
    legend_options = list(
      title = "User Type", 
      style = list(fontSize = "14px")
    ),
    update_view = FALSE # Keep the map centered
  )


```

### Map of Popular End Stations Using Leaflet
```{r}


# Filter to top end stations for each month and user type
top_end_stations <- popular_end_stations[, .SD[which.max(total_rides)], by = .(month, member_casual)]

# Create leaflet map
leaflet(data = top_end_stations) %>%
  addTiles() %>% # Add base map tiles
  addCircleMarkers(
    lat = ~end_lat,
    lng = ~end_lng,
    radius = ~sqrt(total_rides) / 2, # Scale marker size by number of rides
    color = ~ifelse(member_casual == "member", "blue", "red"), # Color by user type
    group = ~member_casual, # Group for filtering
    label = ~paste(
      "Station Name: ", end_station_name, ",",
      "Month: ", month, " ",
      "Total Rides: ", total_rides
    ), # Hover information
    labelOptions = labelOptions(
      style = list(
        "font-weight" = "bold", 
        "font-size" = "12px"
      ),
      textsize = "15px",
      direction = "auto"
    ),
    popup = ~paste0(
      "<b>Station Name:</b> ", end_station_name, "<br>",
      "<b>Total Rides:</b> ", total_rides, "<br>",
      "<b>Month:</b> ", month, "<br>",
      "<b>User Type:</b> ", member_casual
    ) # Clickable information
  ) %>%
  addLayersControl(
    overlayGroups = c("member", "casual"), # Enable filtering by user type
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addLegend(
    position = "bottomright",
    colors = c("blue", "red"),
    labels = c("Member Riders", "Casual Riders"),
    title = "User Type"
  )

```

## 5. Summarize and Recommendations
### Key Findings
- **Usage Trends**: Members take more rides overall but have shorter trip durations on average.
- **Casual Riders**: Ride primarily on weekends and during leisure hours.
- **Members**: Use bikes more consistently across the week, especially during commute hours.

### Recommendations
- Design weekend promotions targeting casual riders.
- Highlight the benefits of annual memberships for regular commuting needs.
- Increase marketing efforts at popular casual rider stations to convert them to members.
