---
title: "Cyclistic User Behavior Analysis (Monthly Data)"
author: "Junior Data Analyst"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load necessary libraries
library(ggplot2)  # For data visualization
library(dplyr)    # For data manipulation
library(lubridate) # For date and time operations
library(data.table)  # For efficient data manipulation
library(leaflet)  # For mapping
```

## 2. Data Preparation
### Load and Combine Monthly Data
```{r}
# List of monthly data file paths
file_paths <- list(
  "divvy-tripdata/202312-divvy-tripdata.csv",
  "divvy-tripdata/202401-divvy-tripdata.csv",
  "divvy-tripdata/202402-divvy-tripdata.csv",
  "divvy-tripdata/202403-divvy-tripdata.csv",
  "divvy-tripdata/202404-divvy-tripdata.csv",
  "divvy-tripdata/202405-divvy-tripdata.csv",
  "divvy-tripdata/202406-divvy-tripdata.csv",
  "divvy-tripdata/202407-divvy-tripdata.csv",
  "divvy-tripdata/202408-divvy-tripdata.csv",
  "divvy-tripdata/202409-divvy-tripdata.csv",
  "divvy-tripdata/202410-divvy-tripdata.csv",
  "divvy-tripdata/202411-divvy-tripdata.csv"
)

# Load and combine data using data.table
monthly_data <- rbindlist(lapply(file_paths, fread))
```

### 2.1 Check and Clean the Data
#### Check for Missing Values
```{r}
# Check for missing values
missing_summary <- colSums(is.na(monthly_data))
print("Missing Values Summary:")
print(missing_summary)
```

#### Remove Missing Values
```{r}
# Remove rows with missing values
monthly_data <- monthly_data[complete.cases(monthly_data)]
```

#### Ensure Proper Date/Time Format
```{r}
# Convert date/time fields to proper format
monthly_data[, started_at := as.POSIXct(started_at, format = "%Y-%m-%d %H:%M:%S")]
monthly_data[, ended_at := as.POSIXct(ended_at, format = "%Y-%m-%d %H:%M:%S")]
```

#### Remove Duplicate Records
```{r}
# Remove duplicate rows
monthly_data <- unique(monthly_data)
```

### 2.2 Add New Calculated Columns
#### Trip Duration
```{r}
# Calculate trip duration in minutes
monthly_data[, trip_duration := as.numeric(difftime(ended_at, started_at, units = "mins"))]
```

#### Day of Week
```{r}
# Extract the day of the week
monthly_data[, day_of_week := weekdays(started_at)]
```

#### Hour of the Day
```{r}
# Extract the hour from the start time
monthly_data[, hour_of_day := as.numeric(format(started_at, "%H"))]
```

## 3. Analyze the Data
### 3.1 Aggregate Metrics by User Type
#### Total Number of Rides by User Type
```{r}
# Total number of rides by user type
total_rides <- table(monthly_data$member_casual)
total_rides
```

#### Average Trip Duration by User Type
```{r}
# Average trip duration by user type
avg_duration <- monthly_data[, .(avg_duration = mean(trip_duration, na.rm = TRUE)), by = member_casual]

# Plot
ggplot(avg_duration, aes(x = member_casual, y = avg_duration, fill = member_casual)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = round(avg_duration, 1)), vjust = -0.5) +
  labs(title = "Average Trip Duration by User Type", x = "User Type", y = "Average Duration (minutes)") +
  theme_minimal()
```

### 3.2 Analyze Usage Patterns
#### Popular Days by User Type
```{r}
# Count rides by day of the week and user type
popular_days <- monthly_data[, .N, by = .(member_casual, day_of_week)]
popular_days$day_of_week <- factor(popular_days$day_of_week, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

# Plot
ggplot(popular_days, aes(day_of_week, N, fill = member_casual)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Rides by Day of the Week and User Type", x = "Day of the Week", y = "Number of Rides", fill = "User Type") +
  theme_minimal()
```

#### Peak Hours by User Type
```{r}
# Aggregate data for heatmap
heatmap_data <- monthly_data[, .N, by = .(hour_of_day, member_casual)]

# Plot
ggplot(heatmap_data, aes(x = hour_of_day, y = member_casual, fill = N)) +
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Peak Hours by User Type", x = "Hour of Day", y = "User Type", fill = "Number of Rides") +
  theme_minimal()
```

## 4. Visualize Insights
### Geographic Maps: Popular Routes and Stations by User Type
```{r}
# Prepare data for mapping
popular_routes <- monthly_data[, .(total_rides = .N), by = .(start_station_name, end_station_name, member_casual, start_lat, start_lng, end_lat, end_lng)]

# Create map with layer control
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(
    data = popular_routes[member_casual == "member"],
    lng = ~start_lng, lat = ~start_lat,
    radius = ~sqrt(total_rides),
    color = "blue",
    group = "Member"
  ) %>%
  addCircleMarkers(
    data = popular_routes[member_casual == "casual"],
    lng = ~start_lng, lat = ~start_lat,
    radius = ~sqrt(total_rides),
    color = "red",
    group = "Casual"
  ) %>%
  addLayersControl(
    overlayGroups = c("Member", "Casual"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  addLegend("bottomright", colors = c("blue", "red"), labels = c("Member", "Casual"), title = "User Type")
```

## 5. Summarize and Recommend
### Key Findings
- **Usage Trends**: Members take more rides overall but have shorter trip durations on average.
- **Casual Riders**: Ride primarily on weekends and during leisure hours.
- **Members**: Use bikes more consistently across the week, especially during commute hours.

### Recommendations
- Design weekend promotions targeting casual riders.
- Highlight the benefits of annual memberships for regular commuting needs.
- Increase marketing efforts at popular casual rider stations to convert them to members.
