---
title: "Tab Panels in R Markdown"
output:
  html_document:
    code_folding: hide
---

# Tables and Graphs in Tabs

## Using `tabset`

### Tables

#### {.tabset}

##### Morning Data
```{r}
library(knitr)

# Create a sample table
morning_data <- data.frame(
  Start_Station = c("Station A", "Station B"),
  Total_Rides = c(120, 90)
)

# Render the table
kable(morning_data, caption = "Morning Rides Table")
```

##### Evening Data
```{r}
# Create another sample table
evening_data <- data.frame(
  Start_Station = c("Station C", "Station D"),
  Total_Rides = c(100, 80)
)

# Render the table
kable(evening_data, caption = "Evening Rides Table")
```

##### Morning Graph
```{r}
library(ggplot2)

# Sample graph for morning data
ggplot(morning_data, aes(x = Start_Station, y = Total_Rides)) +
  geom_col(fill = "blue") +
  ggtitle("Morning Rides")

```

##### Evening Graph
```{r}
# Sample graph for evening data
ggplot(evening_data, aes(x = Start_Station, y = Total_Rides)) +
  geom_col(fill = "orange") +
  ggtitle("Evening Rides")

```


---

### Explanation of the Code:
1. **`tabset`**: Adding `{.tabset}` after a header creates a group of tabs for the subsequent content. Each subheading (`###`) within that tabset becomes a separate tab.
2. **Tables**: Tables are created and displayed under their respective tabs.
3. **Graphs**: Graphs are rendered under their respective tabs using `ggplot2`.

---

### Rendered Output (HTML):
- You’ll see two tabsets:
  - **Tables**: Contains two tabs, "Morning Data" and "Evening Data," displaying the respective tables.
  - **Graphs**: Contains two tabs, "Morning Graph" and "Evening Graph," displaying the respective graphs.

---

### Additional Notes:
1. **Tabsets only work in HTML output.** If you want similar functionality in PDF, consider using LaTeX packages for advanced layout.
2. If you want more advanced tabbed layouts or styling, you can use custom HTML and JavaScript or R packages like `shiny` or `bslib`.

Let me know if you need further customization or help!

